---
title: "Fantasy Football Scores"
author: "FWP Sabermetrics Club"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This project looks at how accurate fantasy football projections can be, and what might be driving that. 
We have data from the 2018 season on weekly projections and actual points earned for different positions and players. 


# Libraries

These are some of the libraries we will need to do our analysis:
```{r}
library(corrplot)
library(skimr)
library(tidyverse) 
```


# Load Data

There are data files for each key position in fantasy football: QB, RB, WR, K, and D/ST. 
Here, we'll read in each file:
```{r}
qbs <- readRDS("../data/fantasy_projections_2018_nfl_QB.Rds")
rbs <- readRDS("../data/fantasy_projections_2018_nfl_RB.Rds")
wrs <- readRDS("../data/fantasy_projections_2018_nfl_WR.Rds")
kick <- readRDS("../data/fantasy_projections_2018_nfl_K.Rds")
dst <- readRDS("../data/fantasy_projections_2018_nfl_DST.Rds")
```



# Running Backs

Note that for some of the stats (`rec_td` or `rush_td`) in the running backs data, when the value should be zero, it is instead reported as missing (in `R`, this is denoted `NA`). 
This means we need to replace some of those "missing" values in the data with 0:
```{r}
rbs <- rbs %>%
  mutate_if(.predicate = function(field) is.numeric(field) & anyNA(field), 
            .funs = function(x) ifelse(is.na(x), 0, x))
```

## Inspect Data

Below are some univariate summaries of the data:

```{r}
rbs %>% skim()
```

Below are some bivariate summaries for the data:
```{r}
rbs %>%
  select(weekpts, site_pts, rush_yds, rush_tds, rec_yds, rec_tds, fumbles_lost, two_pts) %>%
  cor(.) %>%
  corrplot()
```

Finally, below are some plots between variables that might be of interest for our project.

This is plot of projected vs. actual points for running backs:
```{r}
ggplot(rbs) + # start canvas with rb data
  geom_point(aes(site_pts, weekpts)) + # add the points
  stat_function(fun = function(x) x, color = 'red') + # plot line y = x in red
  geom_smooth(aes(site_pts, weekpts), method = "loess") + # add the smoothed loess curve
  labs(title = "Projected vs. Actual Points (Running backs)", 
       x = "Projected Points", 
       y = "Actual Points")
```

# Quarterbacks


# Wide Receivers


# Defense/Special Teams