---
title: "OPOY Predictions"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

These are some of the libraries we will need to do our analysis:
```{r}
library(corrplot)
library(skimr)
library(tidyverse) 
```


# Load Data

We have PFR data on passing, rushing, and receiving stats by season:
```{r}
pass <- readRDS("../data/PFR/pfr_passing_1985_2018.RDS")
rush <- readRDS("../data/PFR/pfr_rushing_1985_2018.RDS")
rec <- readRDS("../data/PFR/pfr_receiving_1985_2018.RDS")
```


We also have a table of OPOY winners:
```{r}
opoy_winners <- readRDS("../data/PFR/pfr_awards_opoy_1972-2018.RDS") %>% filter(Year >= 1985)
```


# Rushing

First, we'll need to clean up the `player` column in the `rush` data:
```{r}
rush$player <- gsub("\\+|\\*", "", rush$player) # clean names

# tidy up names for the winners data
opoy_winners <- opoy_winners %>%
  select(year = Year, 
         player = Player, 
         team = Tm)
```



Now that our names and years are standardized, we can write a quick function to return whether a player won OPOY in a given year or not:
```{r}
rush <- rush %>% 
  mutate(opoy = map2(player, year, 
                     .f = function(x, y){
                       win = nrow(filter(opoy_winners, 
                                   player == x, 
                                   year == y)) == 1
                       return(as.integer(win))
                     }), 
         opoy = unlist(opoy))
```

Finally, we can see from the position field `pos` that this data contains everyone who ever attempted to run in a game. 
Further, some of the positions are a little wonky. 
So, we should filter out folks who only had a few carries (i.e., fewer than 15, 20, or even 50). 

```{r}
rush <- rush %>%
  filter(att > 20) %>% # Note: where do we want to set this?
  mutate(pos = toupper(pos)) # get rid of weird capitalizaiton rules

# There are some odd position alignments that we may need to deal with later
rush %>% distinct(pos)
```


## Inspect Data

```{r}
rush %>% skim()
```


We can get a plot of some important pairwise correlations in our data with the following code:
```{r}
rush %>%
  select(year, gs, yds, td, lng, y_a, y_g, fmb, opoy) %>%
  cor(.) %>%
  corrplot()
```

Not surprisingly, things like the yards and TDs are associated with winning OPOY.


We can also examine whether distributions of variables differ according to whether a player won OPOY or not.
```{r}
ggplot(rush) + 
  geom_density(aes(td, fill = factor(opoy)), alpha = 0.5) + # make sure the variable names are inside the aes() function
  labs(title = "Touchdowns (by Season)", 
       x = "TD", 
       y = "Density")
```


We can also examine whether relationships between variables differ according to whether a player won OPOY or not.
```{r}
ggplot(rush) + # start canvas with rb data
  geom_point(aes(y_g, td, color = factor(opoy)), alpha = 0.6) + # add the points
  geom_smooth(aes(y_g, td, color = factor(opoy)), method = "loess") +  # add the smoothed loess curve
  labs(title = "Yards Per Game vs. TDs", 
       x = "YPG", 
       y = "TD")
```

Note that the pattern seems constant for OPOY winners and losers, but that OPOY winners tended to have more TDs and more yards per game.

NOTE: What other plots should we make?

# Passing

# Receiving