---
title: "Salaries Project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

These are some of the libraries we will need to do our analysis:
```{r}
library(corrplot)
library(skimr)
library(tidyverse) 
```


# Load Data

There are datasets on player salaries by whether they play on defense, offense, or special teams:
```{r}
df <- readRDS("../data/sportrac/defense.RDS")
of <- readRDS("../data/sportrac/offense.RDS")
st <- readRDS("../data/sportrac/special.RDS")
```


# Offense

We might want to add a column that indicates if a player plays a "skill" position:
```{r}
of <- of %>%
  mutate(skill = ifelse(pos %in% c("QB", "RB", "TE", "WR"), 1, 0)) 
```

## Inspect Data

Below are some univariate summaries of the data:

```{r}
of %>% skim()
```


## Plot data

Below are some bivariate summaries for the data:
```{r}
of %>%
  select(yrs, age, dollars, average, guaranteed_at_signing, practical_guarantees, percent_gteed) %>%
  cor(.) %>%
  corrplot()
```

Finally, below are some plots between variables that might be of interest for our project.

For instance, We might also be interested in differences in how much of the contract is guaranteed by position:
```{r}
ggplot(of) + 
  geom_density(aes(percent_gteed, fill = factor(skill)), alpha = 0.5)
```


Or, we can look at stats within skill positions:
```{r}
of %>%
  filter(skill == 1) %>%
  ggplot() +
    geom_density(aes(percent_gteed, fill = pos), alpha = 0.3)
```


# Defense

# Special Teams